Efficient pedigree recording for fast population genetics simulation

### 2019-08-09 talks/poster/probgen-2018-poster.tex
added slides

\newlength{\leftimgwidth}
\begin{poster}%
  % Poster Options
  {
  % Columns and Column spacing
  columns=3,
  colspacing=1.5em,
  % Background
  % background=user,
  background=shade-tb,
  % Poster header options
  eyecatcher=yes,
  posterheaderheight=0.15\textheight,
  % Format of textbox header
  headerborder=open,
  % headershape=small-rounded,
  headershape=roundedright,
  headershade=shade-tb,
  % headershade=plain,
  headerfont=\Large\textsf, %Sans Serif
  % Format of textbox
  boxborder=small-rounded,
  % boxborder=roundedleft,    
  boxshade=plain,
  linewidth=0.5pt,
  % Color style
  % bgColorOne=lightestcyan,
  % bgColorTwo=lightcyan,
  bgColorOne=white,
  bgColorTwo=white,
  % borderColor=cyan,
  borderColor=black,
  headerColorOne=DavisBlue,
  % headerColorOne=cyan,
  headerColorTwo=black,
  % headerFontColor=black,
  headerFontColor=white,
  % boxColorOne=lightercyan,
  boxColorOne=HohenheimLightLightBlue,
  boxColorTwo=white,
  % Show grid to help with alignment
  grid=no
  % grid=yes
  }
  % Eye Catcher
  {
  \makebox[15em][r]{%
    \begin{minipage}{15em}
       \hfill
       \includegraphics[width=15em]{tskit_logo} \\
    \end{minipage}
    }
  }
  % Title
  {\sf %Sans Serif
  %\bf% %bold  
  \vspace{0.5em}
     \textbf{\textcolor{DavisBlue}{Recording pedigrees for fast genome simulation and analysis}}\vspace{0.5em}}
  % Authors
  {\sf %Sans Serif
    Peter Ralph$^{\ddagger}$ \\  \vspace{-1.0mm}
    {\small \textit{$\ddagger$ Mathematics and Biology, University of Oregon} }\\
    {\textbf{code:}  \url{https://github.com/tskit-dev} }\\
    {\textbf{preprint:}  \url{https://www.biorxiv.org/content/early/2018/01/26/248500} }\\
  }
  % Project logo
  {
    \makebox[15em][r]{%
      \begin{minipage}{12em}
        \hfill
          \includegraphics[width=12em]{UOSignature-STK-BLK} \\
      \end{minipage}
    }
  }

@@

For a set of sampled chromosomes,
at each position along the genome there is a genealogical tree
that says how they are related.

A \textbf{tree sequence} describes this sequence of trees.

@@

    \item The \emph{pedigree} (parental relationships) plus crossover locations
        would give us the tree sequence for \emph{everyone, ever}.

@@

    \item This information is equivalent to the Ancestral Recombination Graph (ARG).

@@

\citep{kelleher2016efficient}
introduced the \textbf{tree sequence} data structure
for a fast coalescent simulator, \texttt{msprime}.

@@

@@

        \item \textbf{Edges:}   Who inherits from who; only \emph{necessary} for coalescent events.
            \emph{Records:} interval (left, right); parent node; child node.

        \item \textbf{Nodes:}   The ancestors those happen in.
            \emph{Records:} time ago (of birth); ID (implicit).

        \item \textbf{Mutations:}   When state changes along the tree.
            \emph{Records:} site it occurred at; node it occurred in; derived state.

        \item \textbf{Sites:}   Where mutations fall on the genome.
            \emph{Records:} genomic position; ancestral (root) state; ID (implicit).

@@

    \begin{center}
        \includegraphics[width=0.75\textwidth]{../file_size}
    \end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Contributions}{name=contrib,column=1,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{itemize}
        \compresslist
    \item Jerome Kelleher -- algorithms, data structures, and core development
    \item Kevin Thornton -- implementation, profiling with \texttt{fwdpp},
    \item Jaime Ashander -- implementation, profiling with \texttt{simuPOP}
    \item Ben Haller -- implementation in \texttt{SLiM}
    \item Jared Galloway -- implementation in \texttt{SLiM}
    \end{itemize}

@@

\includegraphics[width=7em]{../finger_right} {\Large \emph{This may be a bad idea.}}

@@

        \item add each gamete to the Node Table,
        \item add entries to the Edge Table
            recording which parent each gamete inherited each bit of genome from, and
        \item add any new selected mutations to the Mutation Table 
            and (if necessary) their locations to the Site Table.

@@

\textbf{But,} we won't end up needing the entire history of everyone ever.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Simulations, $50\times$ faster}{name=fast,column=2,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    Recording, simplifying, and output of tables, was done with \texttt{C} code in \texttt{tskit}.
    The simulation was done in \texttt{fwdpp} (C++ by Kevin Thornton),
    connected with \texttt{pybind11} and \texttt{numpy}.
    \textit{Machine:} Ubuntu / 2x 2.6 GHz Intel E5-2650 CPU.

    \emph{Simulation parameters:}
    \begin{itemize}
            \compresslist
        \item  Wright-Fisher population of size $N$, for $10N$ generations
        \item  neutral mutation rate $\mu$ equal to recombination rate $r$ per gamete
        \item  many, weakly deleterious mutations: rate $\mu/100$ with
            $s$ exponentially distributed with mean $2.5/N$.
    \end{itemize}

    \emph{Note:}
    simulations that recorded tree sequences ("pedigree recording")
    had neutral mutation rate was zero,
    but neutral mutations were added \emph{afterwards}.

    \begin{center}
    \includegraphics[width=\textwidth]{rawspeed}

    \includegraphics[width=0.75\textwidth]{speedup}
    \end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Fast statistics computation}{name=stats,column=2,below=fast}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \textbf{A general class of statistics:}
    any polynomial $f(p_1, \ldots, p_k)$ gives a statistic on $k$ populations,
    by setting $p_i$ to the allele frequency in population $i$ and averaging $f$ across loci.
    \emph{(Example: divergence is $p_1(1-p-2) + p-2(1-p_1)$.)}

    The polynomial defines a weighting function on edges of the trees in the tree sequence,
    and the statistic can be calculated by summing across mutations
    weighted by the edge they occur on.
    \emph{(Example: proportion of pairs from the two populations on opposite sides of the edge.)}

    \textbf{Computation on a tree sequence:}
    If we compute weights on the first tree, then we only need update those weights above nodes that change
    in the second tree.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{}{name=references,column=2,below=summary}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

  \scriptsize
  \renewcommand{\section}[2]{\vskip 0.0em}
  \bibliographystyle{abbrvnat}
  \setlength{\bibsep}{0.0pt}
  \bibliography{../../references}
  }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\newlength{\leftimgwidth}
\begin{poster}%
  % Poster Options
  {
  % Columns and Column spacing
  columns=3,
  colspacing=1.5em,
  % Background
  % background=user,
  background=shade-tb,
  % Poster header options
  eyecatcher=yes,
  posterheaderheight=0.15\textheight,
  % Format of textbox header
  headerborder=open,
  % headershape=small-rounded,
  headershape=roundedright,
  headershade=shade-tb,
  % headershade=plain,
  headerfont=\Large\textsf, %Sans Serif
  % Format of textbox
  boxborder=small-rounded,
  % boxborder=roundedleft,    
  boxshade=plain,
  linewidth=0.5pt,
  % Color style
  % bgColorOne=lightestcyan,
  % bgColorTwo=lightcyan,
  bgColorOne=white,
  bgColorTwo=white,
  % borderColor=cyan,
  borderColor=black,
  headerColorOne=DavisBlue,
  % headerColorOne=cyan,
  headerColorTwo=black,
  % headerFontColor=black,
  headerFontColor=white,
  % boxColorOne=lightercyan,
  boxColorOne=HohenheimLightLightBlue,
  boxColorTwo=white,
  % Show grid to help with alignment
  grid=no
  % grid=yes
  }
  % Eye Catcher
  {
  \makebox[15em][r]{%
    \begin{minipage}{15em}
       \hfill
       \includegraphics[width=15em]{tskit_logo} \\
    \end{minipage}
    }
  }
  % Title
  {\sf %Sans Serif
  %\bf% %bold  
  \vspace{0.5em}
     \textbf{\textcolor{DavisBlue}{Tree sequences \\ for fast genome simulation and analysis}}\vspace{0.5em}}
  % Authors
  {\sf %Sans Serif
    Peter Ralph$^{\ddagger}$ \\  \vspace{-1.0mm}
    {\small \textit{$\ddagger$ Mathematics and Biology, University of Oregon} }\\
    {\textbf{code:}  \url{https://github.com/tskit-dev} }\\
    \vspace{0.8em}
    {Related posters: Jerome Kelleher (tree sequences) and Phillip Messer (SLiM)}
    % {\textbf{paper:}  \url{https://www.biorxiv.org/content/early/2018/01/26/248500} }\\
  }
  % Project logo
  {
    \makebox[15em][r]{%
      \begin{minipage}{12em}
        \hfill
          \includegraphics[width=10em]{UOSignature-STK-BLK} \\
      \end{minipage}
    }
  }

@@

A \textbf{tree sequence} describes
the correlated sequence of genealogical trees
that say how a set of chromosomes are related.

@@

    \item The \emph{pedigree} plus crossover locations
        would give us the tree sequence for \emph{everyone, ever}.

@@

    \item This is equivalent to the Ancestral Recombination Graph (ARG).

@@

\citet{kelleher2016efficient}
introduced the \textbf{tree sequence} data structure
for a fast coalescent simulator, \texttt{msprime}.
Updated in \citet{kelleher2018efficient}.

@@

    % We updated and generalized the data structure in \citet{kelleher2018efficient}:

@@

        \item \textbf{Edges:}   Who inherits from who. \\
            \emph{Records:} interval (left, right); parent node; child node.

        \item \textbf{Nodes:}   The ancestors those happen in. \\
            \emph{Records:} time ago (of birth); individual.

        \item \textbf{Mutations:}   When state changes along the tree. \\
            \emph{Records:} site index; node index; derived state.

        \item \textbf{Sites:}   Where mutations fall on the genome. \\
            \emph{Records:} genomic position; root state.

        \item \textbf{Individuals:} Optional. Containers for polyploids. \\
            \emph{Records:} metadata; pointed to by nodes.

@@

    \begin{center}
        \includegraphics[width=0.75\textwidth]{storing_everyone}
    \end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{References}{name=contrib,column=1,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  % \scriptsize
  \renewcommand{\section}[2]{\vskip 0.0em}
  \bibliographystyle{abbrvnat}
  \setlength{\bibsep}{0.0pt}
  \bibliography{refs}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

        \item Add new individuals and their chromosomes to the Individual and Node Tables, respectively.
        \item Record in the Edge Table 
            which parental chromosome each gamete inherited each bit of genome from.
        \item Add any new selected mutations to the Mutation Table 
            and (if necessary) their locations to the Site Table.

@@

\includegraphics[width=7em]{../finger_right} 
\parbox[bottom]{20em}{
    \vspace{-4em}
    {\Large \emph{This is too much data.}} \\
    \emph{\ldots but,} most of it can be \emph{simplified} away.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Simulations, $50\times$ faster}{name=fast,column=1,below=howfwd}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    Recording, simplifying, and output of tables, is done with \texttt{C} code in \texttt{tskit}.

    % The simulation was done in \texttt{fwdpp} (C++ by Kevin Thornton),
    % connected with \texttt{pybind11} and \texttt{numpy}.
    % \textit{Machine:} Ubuntu / 2x 2.6 GHz Intel E5-2650 CPU.

    % \emph{Simulation parameters:}
    % \begin{itemize}
    %         \compresslist
    %     \item  Wright-Fisher population of size $N$, for $10N$ generations
    %     \item  neutral mutation rate $\mu$ equal to recombination rate $r$ per gamete
    %     \item  many, weakly deleterious mutations: rate $\mu/100$ with
    %         $s$ exponentially distributed with mean $2.5/N$.
    % \end{itemize}

    \begin{center}
    % \includegraphics[width=\textwidth]{rawspeed}

    \parbox[bottom]{0.49\textwidth}{
        \includegraphics[width=0.45\textwidth]{speedup}
        (above) many selected mutations;
        (right) neutral.
    }
    \parbox[bottom]{0.49\textwidth}{
        \includegraphics[width=0.45\textwidth]{slim_speed}
    }
    \end{center}

    Simulations that recorded tree sequences
    had neutral mutation rate set to zero during the simulation,
    and neutral mutations added \emph{afterwards}.

    \vspace{2em}

    Results in \citet{haller2018treesequence,kelleher2018efficient}.

    \vspace{2em}

    {\Large
    Now in SLiM and fwdpp!
    }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{A general class of statistics}{name=stats,column=2,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

        Every single-site statistic is a function from genotype patterns to $\mathbb{R}$,
        and each SNP genotype pattern is determined by 
            the samples below the edge it occurs on.
            \vspace{1em}

    \emph{Ingredients:}
    \begin{enumerate}
            \compresslist
        \item A list of $k$ subsets of samples, $I_1, \ldots, I_k$,
            with $|I_j| = N_j$.
        \item A weighting function $f : \mathbb{N}^k \to \mathbb{R}$.
    \end{enumerate}

    \emph{Computation} on a genome of length $L$ with with $M$ SNPs:
    \begin{enumerate}
            \compresslist
        \item Let $n_i(j)$ be the number of samples in $I_j$
            that inherit mutation $i$.
        \item Use the weighting function to compute the weight of each mutation;
            average these across the genome:
            \[
                S(w) := \dfrac{1}{L} \sum_{i=1}^M (w(n_1, \ldots, n_k)  
                                        + w(N_1 - n_1, \ldots, N_k - n_k) .
            \]
    \end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Examples}{name=examples,column=2,below=stats}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    Folded SFS:
        $$ f_\ell(n_1) = \textbf{1}_\ell(n_1) \qquad \text{for}\; 1 \le \ell \le N_1 - 1 .$$

    Divergence between $I_1$ and $I_2$:
        $$ w(n_1, n_2) = n_1 (N_2 - n_2) / N_1 N_2 .$$

    Patterson's $f_4(I_1, I_2; I_3, I_4)$:
        \begin{align*}
           &w(n_1, n_2, n_3, n_4) \\
            & \qquad =
            \frac{
               n_1 n_3 (N_2 - n_2) (N_4 - n_4)
                      - n_1 n_4 (N_2 - n_2) (N_3 - n_3) }{ N_1 N_2 N_3 N_4  }
              .  \end{align*}

}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Computation}{name=computation,column=2,below=examples}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{enumerate}
            \compresslist
        \item[0.]
            Find $n$ for each node in the first tree.
        \item[1.]
            Compute $f(n)$ for each mutation on this tree,
            and add these to the total.
        \item[2.]
            Update the tree, and update $n$ for each node
            in the path from changed nodes to the root.
        \item[3.]
            Return to (1).
        \item[4.]
            When done, divide the total by the sequence length.
    \end{enumerate}

    \emph{Complexity} for $N$ samples at $L$ SNPs with $T$ trees:
        $$ O(N + L + T \log(N)) . $$
    This is much better than using the genotype matrix, which is $O(NL)$.

}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Other notes}{name=other,column=2,below=computation}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{itemize}
        \item 
            The \emph{same} statistics can be computed 
            as weighted averages of \textbf{branch lengths} also,
            which gives the expected value of the statistic, averaged over mutational noise.

        \item 
            This tracks who is ancestral to a set of samples;
            \emph{a dual procedure} tracks who has inherited from a set of ancestors.

        \item
            \emph{Many statistics} can be computed on the same pass through the tree sequence
            with little additional cost.

        \item
            The implementation computes these in \emph{windows}.

    \end{itemize}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Updating internal state}{name=statfig,column=2,below=other}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \includegraphics[width=0.3\textwidth]{stats_1}
    \includegraphics[width=0.3\textwidth]{stats_2}
    \includegraphics[width=0.3\textwidth]{stats_3}

}

@@

@@
